---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


In this assignment, I will answer several questions about a set of walking activity data for an anonymous individual, described in the README file.  

## Loading and preprocessing the data

Our first step is to load libraries, the data and do any initial preprocessing of the data.  

First, let's load the `dplyr` package, which provides convenient methods for grouping, filtering, and summarizing data.  We'll also load `ggplot2` for attractive plots.  


```{r message=FALSE}
#message=FALSE in the code chunk header just prevents R masking messages due to loading dplyr

library(dplyr)      #for easier data summaries
library(ggplot2)    #for nice looking plots

```

Next, let's read in the data file to a data frame called **activity**.

```{r}
# tbl_df is a dplyr wrapper around a data frame 
# that makes it print only a few rows and only the 
# colums that will fit on the screen.

activity<-tbl_df(read.csv(file="activity.csv", head=TRUE))
activity
```

As described in the README file, there are 17,568 rows and 3 variables - steps, date, and interval. 

Let's now create a version of the data without the rows that have an NA value for steps.  This is not the same thing as the data imputation we will do later.  We're removing the NAs here because making summaries and plots without this handling of NAs doesn't seem useful.  Otherwise, if you just use `na.rm=TRUE` you get a 0 for your summary statistic on a day that has all NAs for steps, which might not be a good characterization of the typical daily activity.  

We'll use the `dplry` `filter` functionality to only select rows where the value of steps is not NA. 

```{r}
activity_no_NA<-filter(activity, !(is.na(steps)))
activity_no_NA

```

## What is mean total number of steps taken per day?

Let's look at steps per day to get a sense of the distribution of daily activity for this person.  Again, we'll use `dplyr` to help us do the summary.  Then we'll use `ggplot2` to create the histogram.  

First, let's summarize total steps by day.  We'll do this by treating date as a factor and by using the **activity_no_NA** data set, from which NAs were removed.  I use the `dplyr` pipeline notation `%>%`, which improves code readability for a series of related programming statements.  

Our 2 summary statistics are the **total** steps for a particular day, and the **count** of (non-NA) observations for that particular day.  


```{r}
stepsByDay<- activity_no_NA %>%
  group_by(Date = as.factor(date)) %>%
  summarise(total = sum(steps), count = n())

head(stepsByDay,10)
```

I can now create a histogram, using `ggplot2`, of daily steps.  I use a binwidth of 2000 in order to give a good balance between summarizing the data and still seeing enough detail.


```{r StepsHistogram}
stepHisto<- ggplot(stepsByDay, aes(x=total)) +
  geom_histogram(binwidth=2000,color="black", fill="white") + 
  labs(title="Total Daily Steps\n ") +                         
  labs(x = "\nTotal Steps") +
  labs(y = "Number of Days\n")

stepHisto

```


We'll use the `summary` function from base R to calculate the **mean**, **median**, quartiles and extremes for the distribution shown in the histogram above.  I can see that the **mean and the median are just above 10,000 steps**.  Specifically, the **mean=10,770** and the **median=10,760**.


```{r}
#get the mean, median, etc. from the 'summary' function in base R.
#We create our own array, DailySteps, by taking the different elements from the summary output.

stepsSummary<-summary(stepsByDay$total) 
DailySteps<-c(                         
  as.integer(stepsSummary[4]), #mean
  as.integer(stepsSummary[3]), #median
  as.integer(stepsSummary[1]), #min
  as.integer(stepsSummary[6]), #max
  as.integer(stepsSummary[2]), #1stQuartile
  as.integer(stepsSummary[5])  #3rdQuartile
  )

#We assign names to the array I just created.
names(DailySteps)<-c("Mean", "Median", "Minimum", "Maximum", "1Q", "3Q")            
#Make it a data frame so it prints vertically, and maybe I can do more with it later.
DailySteps<-as.data.frame(DailySteps)

DailySteps
```

## What is the average daily activity pattern?

Let's begin by looking at a plot of the average number of steps per 5-minute period averaged over the 53 days (the ones without NAs) we've been analyzing thus far.  Only those times that mark the beginning of an interesting change in the pattern are marked on the x-axis in order to keep the plot cleaner.  Also, our goal is to get a general idea of the shape of the daily activity pattern.  More detail could be seen by inspecting the data in the **stepsByInterval** data frame.

First, we'll average the data by interval.

```{r}
stepsByInterval<- activity_no_NA %>%
  group_by(Interval = as.factor(interval)) %>%
  summarise(AverageSteps = mean(steps))

head(stepsByInterval,5)
```

Then, I will use the **stepsByInterval** data frame to create a plot of the average steps by interval.

```{r stepsByIntervalPlot}
#We will only label some x-axis tickmarks to keep the plot cleaner.

#define which intervals you want to label on the x-axis.
#these were identified by inspecting the stepsByInterval data frame.
tickmarks<-c(1,68,104,145,187,235) 
#create human readable labels for the tickmarks.
intervalLabels<-c("midnight","5:35","8:35","12:00","15:30","19:30")

#Use the base plotting system to create a time-series plot.
#I find the base plotting system is easier to use for time-series than ggplot2.
plot(stepsByInterval$AverageSteps, type="l", xaxt='n', 
     ylab="Average Steps", xlab="5-minute Interval",main="Average Steps per 5-minute Interval")

#put the tickmarks on the axis with the appropriate labels
axis(1, at=tickmarks, labels = intervalLabels)
```

The **busiest 5-minute interval of the day starts at 8:35am** with an average of just over 206 steps.  The code below produces that result.

```{r}
#Find the interval with the maximum number of steps.
filter(stepsByInterval, AverageSteps == max(AverageSteps))
```

## Imputing missing values

### Total number of missing values in the dataset

There were 8 days in which all 288 observations were NA and no days had fewer NAs than 288, as demonstrated by the following code chunk.  

```{r}
#We're starting with the activity data frame, which is the original data with all the NAs.
TotalNAs<- activity %>%
  group_by(Date = as.factor(date)) %>%
  summarise(TotalNAs = sum(is.na(steps)))

filter(TotalNAs, TotalNAs>0) #show the days where the number of NAs > 0
```

The **total number of NAs in the dataset** therefore is 8*288=2304.  This can be independently verified by counting the total without regard to grouping them by date as a factor.

```{r}
TotalNAs2<- activity %>%
  summarise(TotalNAs = sum(is.na(steps)))

TotalNAs2
```

### Strategy for handling NAs

The best strategy will be to use the mean for the 5-minute interval, as suggested in the assignment.  We've already calculated these means earlier and stored them in the **stepsByInterval** data frame.  Also, we have already created a data frame that has no NAs called **activity_no_NA**.  We also confirmed that the NAs were all within specific days.  There were no days that had some NAs and some data values.  So, our strategy for creating the data set with imputed values will be to create a data frame with just NAs, replace those NAs with the corresponding 5-minute interval mean,  then combine that with the **activity_no_NA** data.

### New dataset from NA strategy.

First we'll create the data that only contains NAs.  There should be 8*288=2304 rows.

```{r NAs}
activity_all_NA<-filter(activity, (is.na(steps)))
activity_all_NA
```

Next we replace the NA data with the corresponding 5-minute interval data.  We'll take advantage of the fact that R recycles values.

```{r impute}
#Choose rows that have NA values for steps
activity_all_NA<-filter(activity, (is.na(steps)))
#Put the average steps per 5-minute interval, stored in stepsByInterval, 
#into the corresponding NA value.  This takes advantage of R's recycling property.
activity_all_NA$steps<-stepsByInterval$AverageSteps
activity_all_NA
```

Finally, we combine the imputed data with the no-NA data.  We can see that we have 17,568 rows, just like the original data set.

```{r combine}
#we use the dplyr efficient rbind function.
imputedActivityData<-rbind_list(activity_all_NA,activity_no_NA)
imputedActivityData
```

### Histogram

We'll re-use our histogram code, but with the imputed data.  

```{r StepsHistogram2}
#We need to compute stepsByDayImputed to feed the histogram
stepsByDayImputed<- imputedActivityData %>%
  group_by(Date = as.factor(date)) %>%
  summarise(total = sum(steps), count = n())

#Plot the histogram
stepHisto2<- ggplot(stepsByDayImputed, aes(x=total)) +
  geom_histogram(binwidth=2000,color="black", fill="white") + 
  labs(title="Total Daily Steps with Imputed Data\n ") +                         
  labs(x = "\nTotal Steps") +
  labs(y = "Number of Days\n")

stepHisto2

```

The histograms look the same.  Let's compare the mean, median, quartiles and extremes for the imputed and non-imputed data.

```{r}
#as we did earlier, use the Base R summary function.
stepsSummaryImputed<-summary(stepsByDayImputed$total) 
DailyStepsImputed<-c(                         
  as.integer(stepsSummaryImputed[4]), #mean
  as.integer(stepsSummaryImputed[3]), #median
  as.integer(stepsSummaryImputed[1]), #min
  as.integer(stepsSummaryImputed[6]), #max
  as.integer(stepsSummaryImputed[2]), #1stQuartile
  as.integer(stepsSummaryImputed[5])  #3rdQuartile
  )

#We assign names to the array I just created.
names(DailyStepsImputed)<-c("Mean", "Median", "Minimum", "Maximum", "1Q", "3Q")  

#Make it a data frame so it prints vertically.
DailyStepsImputed<-as.data.frame(DailyStepsImputed)
#combine the non-imputed summary and the imputed summary.
CompareDailySteps<-cbind(DailyStepsImputed,DailySteps)

CompareDailySteps
```

The data show differences in the quartiles (1st, 3rd, and median).  The mean, min, and max are the same.  Because we imputed mean values, the means weren't different from the data where NAs were removed.  But, the total N differed between the two data sets, which will affect the quartiles.

## Are there differences in activity patterns between weekdays and weekends?
