s---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


In this assignment, we will answer several questions about a set of walking activity data for an anonymous individual, described in the README file.

## Loading and preprocessing the data

Our first step is to load the data and do any initial preprocessing.  

First, let's load the `dplyr` package, which provides convenient methods for grouping, filtering, and summarizing data.  Depending on the configuration of your system, you may see several system messages related to loading `dplyr`.  We'll also load `ggplot2` so that we can make attractive plots (though we could have also used the base plotting system or lattice).  


```{r}
library(dplyr)      #for easier data summaries
library(ggplot2)    #for nice looking plots

```

Next, let's read in the data file.  `tbl_df` is a `dplyr` wrapper around a data frame that makes it print only a few rows and only the colums that will fit on the screen.

```{r}
activity<-tbl_df(read.csv(file="activity.csv", head=TRUE))
activity
```

As described in the README file, there are 17,568 rows and 3 variables - steps, date, and interval. 

Let's now create a version of the data with the rows that have an NA value for steps removed.  We will use that when summarizing steps by day.  Otherwise, if you just use `na.rm=TRUE` you get a 0 for your summary statistic on a day that has all NAs for steps, which might not be a good characterization of the typical daily activity.  We'll use the `dplry` `filter` functionality to only select rows where the value of steps is not NA. 

```{r}
activity_no_NA<-filter(activity, !(is.na(steps)))
activity_no_NA

```

## What is mean total number of steps taken per day?

Let's look at steps per day to get a sense of the distribution of daily activity for this person.  Again, we'll use `dplyr` to help us do the summary.  Then we'll use `ggplot2` to create the histogram.  

First, let's summarize total steps by day.  We'll do this by treating date as a factor and by using the **activity_no_NA** data set, from which NAs were removed.  We use the `dplyr` pipeline notation `%>%`, which improves code readability for a series of related programming statements.  

Our 2 summary statistics are the **total** steps for a particular day, and the **count** of (non-NA) observations for that particular day.  


```{r}
stepsByDay<- activity_no_NA %>%
  group_by(Date = as.factor(date)) %>%
  summarise(total = sum(steps), count = n())

head(stepsByDay,10)
```

The data below show days where the total number of NAs > 0.  We see that there were 8 days in which every observation was an NA and no days had fewer NAs than that.  We will use this fact the basis for our **data imputation** strategy later.


```{r}
TotalNAs<- activity %>%
  group_by(Date = as.factor(date)) %>%
  summarise(TotalNAs = sum(is.na(steps)))

filter(TotalNAs, TotalNAs>0) #use the dplyr filter function to show the rows where the number of NAs > 0
```

We can now create a histogram, using `ggplot2`, of daily steps for the 53 days for which data were collected.  We use a binwidth of 2000 in order to give a good balance between summarizing the data and still seeing enough detail.


```{r StepsHistogram}
stepHisto<- ggplot(stepsByDay, aes(x=total)) +
  geom_histogram(binwidth=2000,color="black", fill="white") + 
  labs(title="Total Daily Steps\n ") +                         
  labs(x = "\nTotal Steps") +
  labs(y = "Number of Days\n")

stepHisto


```


We'll use the `summary` function from base R to calculate the **mean**, **median**, quartiles and extremes for the distribution shown in the histogram above.  We can see that the **mean and the median are just above 10,000 steps**.


```{r}
#get the mean, median, etc. from the 'summary' function in base R.
#We create our own array, DailySteps, by taking the different elements from the summary output.

stepsSummary<-summary(stepsByDay$total) 
DailySteps<-c(                         
  as.integer(stepsSummary[4]), #mean
  as.integer(stepsSummary[3]), #median
  as.integer(stepsSummary[1]), #min
  as.integer(stepsSummary[6]), #max
  as.integer(stepsSummary[2]), #1stQuartile
  as.integer(stepsSummary[5])  #3rdQuartile
  )

#We assign names to the array we just created.
names(DailySteps)<-c("Mean", "Median", "Minimum", "Maximum", "1Q", "3Q")            
#Make it a data frame so it prints vertically, and maybe we can do more with it later.
DailySteps<-as.data.frame(DailySteps)

DailySteps
```

## What is the average daily activity pattern?

Let's begin by looking at a plot of the average number of steps per 5-minute period averaged over the 53 days we've been analyzing thus far.  Only those times that mark the beginning of an interesting change in the pattern are marked on the x-axis in order to keep the plot cleaner.  Also, our goal is to get a general idea of the shape of the daily activity pattern.  More detail could be seen by inspecting the data in the **stepsByInterval** data frame.

First, we'll average the data by interval.  We'll use the data frame that has the NAs removed.

```{r}
stepsByInterval<- activity_no_NA %>%
  group_by(Interval = as.factor(interval)) %>%
  summarise(AverageSteps = mean(steps))
```

Then, we will use the stepsByInterval data frame to create a plot of the average steps by interval.

```{r stepsByIntervalPlot}
#We will only label some x-axis tickmarks to keep the plot cleaner.

#define which intervals you want to label on the x-axis
tickmarks<-c(1,68,104,145,187,235) 
#create human readable lables for the tickmarks.
intervalLabels<-c("midnight","5:35","8:35","12:00","15:30","19:30")

#Use the base plotting system to create a time-series plot.
plot(stepsByInterval$AverageSteps, type="l", xaxt='n', 
     ylab="Average Steps", xlab="5-minute Interval",main="Average Steps per 5-minute Interval")

#put the tickmarks on the axis with the appropriate labels
axis(1, at=tickmarks, labels = intervalLabels)
```

The **busiest 5-minute interval of the day starts at 8:35am** with an average of just over 206 steps.  The code below produces that result.

```{r}
#Find the interval with the maximum number of steps.
filter(stepsByInterval, AverageSteps == max(AverageSteps))
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
